---
title: "Week 1 Methods: Growth"
subtitle: "ECON 125: The Science of Population"
fontsize: 9pt
output: html_document
---

## Setup
Today, we analyze population growth and its components in this dataset:

  - Population, births, deaths, net migration (in 1000s)
  
  - One row per year since 1950 for every country
  
  - From United Nations World Population Prospects

Start by setting up R, loading the dataset, and assigning it a name using `<-`
```{r, message = FALSE}
# Load tidyverse and clear the R environment
library(tidyverse)
rm(list=ls())

# Load dataset and call it country_year_df
country_year_df <- read_csv(url("https://github.com/tomvogl/econ125/raw/main/data/UN_pop_change_country_year.csv"))

# Ask R to only report 2 significant digits
# This is not important and just makes the output easier to read
options(digits = 2)
```

## Variables

You can see which variables are in the dataset by...

 - Clicking on the data frame in the Environment pane in RStudio
 
 - Using the `names()` function in your code
 
```{r}
names(country_year_df)
```

## Data structure

Let's look at the first 10 rows of the dataset

You can see that each country-year gets its own row

  - In the 1950s, Burundi had 2-3 million people

  - Each year, 120k were born, 55k died, and 15k more left than entered

```{r}
head(country_year_df, 10)
```

## Country-level to world-level
Let's study the world population by summing across countries in each year

We'll use three tools from `tidyverse`: the pipe `|>`, `group_by()`, `summarise()`

```{r}
world_year_df <-
  country_year_df |>
  group_by(year) |> 
  summarise(world_pop = sum(pop),
            world_births = sum(births),
            world_deaths = sum(deaths),
            world_netmigration = sum(netmigration))
```

Line-by-line explanation:

  1. Create a new data frame called `world_year_df`
  2. Start with our existing data frame `country_year_df`, and then...
  3. Group the data by `year`, and then...
  4. Within each year, sum each variable across countries
  
## New data frame
Here are the first 10 rows of the new data frame we created

  - Net migration is always zero (slight nonzero values due to rounding errors)
  
  - Aliens did not land on earth, and Elon did not start his Mars colony
```{r}
head(world_year_df, 10)
```

## Summary statistics
R offers many ways to calculate summary statistics for a variable

  - We will use `summarise()` from `tidyverse`, along with the functions 
  `mean()`, `sd()`, `min()`, and `max()` 
  
  - The `summary()` function from base R is also useful, but its syntax is 
  different from `tidyverse`, so we will skip it to avoid confusion
  
Here is an example for net migration, which again confirms that it is always zero:
```{r}
world_year_df |> summarise(mean = mean(world_netmigration),
                           std_dev = sd(world_netmigration),
                           minimum = min(world_netmigration),
                           maximum = max(world_netmigration))
```

## Time series of the world population
Now let's plot the world population over time

We'll use the `tidyverse` tool for drawing graphs: `ggplot()`

```{r, fig.width=5, fig.height=2.5}
ggplot(world_year_df, aes(x = year, y = world_pop)) +
  geom_line() +
  scale_y_continuous("Population (in 1000s)") + 
  scale_x_continuous("Year") 
```

## Levels to growth rates
It's hard to see details in the time series of population

Let's calculate the growth rate and add it to `world_year_df`

```{r}
world_year_df <- 
  world_year_df |>
  arrange(year) |>
  mutate(world_popgrowth = 
           100*(lead(world_pop) - world_pop)/world_pop)
```

Some comments:

  - `arrange()` sorted the data by year
  
  - `mutate()` generated a new variable `world_popgrowth`
  
  - `lead()` looked forward to the next value of `world_pop`
  
  - The line breaks were not necessary; they just made my slides prettier

## Plotting the population growth rate over time
When we plot the growth rate, the results are more interesting!

  - World population growth steadily declining for decades
  - Large spike down in 1960, due to the Great Famine in China

```{r, fig.width=5, fig.height=2.5, warning = FALSE}
ggplot(world_year_df, aes(x = year, y = world_popgrowth)) +
  geom_line()+
  scale_y_continuous("Population growth rate") + 
  scale_x_continuous("Year") 
```

## Decomposing population growth
Recall the demographic balancing equation:
$$P_1 = P_0 + (B-D) + (I-E)$$
We can rearrange as follows:
$$\frac{P_1-P_0}{P_0} = \frac{B}{P_0} - \frac{D}{P_0} + \frac{I-E}{P_0}$$
The population growth rate equals the birth rate minus the death rate plus the net migration rate. Since global net migration is 0, focus on births and deaths:
```{r}
world_year_df <- 
  world_year_df |>
  arrange(year) |>
  mutate(world_birthrate = 100*world_births/world_pop, 
         world_deathrate = -100*world_deaths/world_pop) 
```

## Time series of population growth rate, birth rate, and death rate
To plot all three series, we modify the `ggplot()` syntax as follows:
```{r, fig.width=5, fig.height=2.5, warning = FALSE}
ggplot(world_year_df, aes(x = year)) +
  geom_line(aes(y = world_popgrowth, color = "Growth")) +
  geom_line(aes(y = world_deathrate, color = "Death")) +
  geom_line(aes(y = world_birthrate, color = "Birth")) +
  scale_y_continuous("Rate") + 
  scale_x_continuous("Year")
```

## Country case study
I claimed that Great Chinese Famine accounted for the blip in 1960. Let's check.

To focus on China, we use `filter()`:
```{r}
china_df <-
  country_year_df |>
  filter(country=="China") |>
  arrange(year) |>
  mutate(popgrowth = 100*(lead(pop) - pop)/pop,
         birthrate = 100*births/pop, 
         deathrate = -100*deaths/pop,
         migrate = 100*netmigration/pop)
```

## Population growth and its components in China
```{r, fig.width=5, fig.height=2.5, warning = FALSE}
ggplot(china_df, aes(x = year)) +
  geom_line(aes(y = popgrowth, color = "Growth")) +
  geom_line(aes(y = deathrate, color = "Death")) +
  geom_line(aes(y = birthrate, color = "Birth")) +
  geom_line(aes(y = migrate, color = "Migration")) +
  scale_y_continuous("Rate") + 
  scale_x_continuous("Year")
```

## Chinese demographic history
Some interesting facts:

  1. Death rates rose **and** birth rates fell during Great Leap Forward (1958-62)
  
  2. Most decline in the birth rate happened **before** the One Child Policy (1979)
  
  3. After end of One Child Policy (2016), fertility fell more
  
  4. Large reductions in death rate before 1980, some slippage recently

```{r, fig.width=5, fig.height=2.5, warning = FALSE, echo = FALSE}
ggplot(china_df, aes(x = year)) +
  geom_line(aes(y = popgrowth, color = "Growth")) +
  geom_line(aes(y = deathrate, color = "Death")) +
  geom_line(aes(y = birthrate, color = "Birth")) +
  geom_line(aes(y = migrate, color = "Migration")) +
  scale_y_continuous("Rate") + 
  scale_x_continuous("Year")
```

## Setting up country comparisons
Let's compare the five countries with the largest populations in 1950

  - The symbol `|` means **or**
  
  - `group_by()` and `ungroup()` make sure we take leads in the same country
  
```{r}
top5_df <- 
  country_year_df |>
  filter(country == "China" | 
         country == "India" |
         country == "United States of America" |
         country == "Russian Federation" |
         country == "Japan") |>
  group_by(country) |>
  arrange(year) |>
  mutate(popgrowth = 100*(lead(pop) - pop)/pop,
         birthrate = 100*births/pop, 
         deathrate = -100*deaths/pop,
         migrate = 100*netmigration/pop) |>
  ungroup()
```

## Shortening country names
Some country names are very long!

Use `unique()` to list the unique values of the variable `country` in the data frame `top5_df`
```{r}
unique(top5_df$country)
```

Let's shorten the longest ones using `mutate()` and `case_when()`
```{r}
top5_df <- 
  top5_df |>
  mutate(
  country = case_when(country=="United States of America" ~ "USA",
                      country=="Russian Federation" ~ "Russia",
                      .default = country)
  )
```

## Population growth over time by country

Plot all five countries in one panel using the `color` option:
```{r, fig.width=5, fig.height=2.5, warning = FALSE}
ggplot(top5_df, aes(x = year, y = popgrowth, color = country)) +
  geom_line()+
  scale_y_continuous("Population growth rate") + 
  scale_x_continuous("Year")
```

## Population growth and its components over time by country

Plot one panel per country using `facet_wrap()`:
```{r, fig.width=5.5, fig.height=2.5, warning = FALSE}
ggplot(top5_df, aes(x = year)) +
  geom_line(aes(y = popgrowth, color = "Growth")) +
  geom_line(aes(y = deathrate, color = "Death")) +
  geom_line(aes(y = birthrate, color = "Birth")) +
  geom_line(aes(y = migrate, color = "Migration")) +
  scale_y_continuous("Rate") + 
  scale_x_continuous("Year") +
  facet_wrap(~country)
```