---
title: "Week 2 Methods: Mortality"
subtitle: "ECON 125: The Science of Population"
fontsize: 9pt
output: html_document
---

## Setup

Today, we analyze mortality across countries. Our data are:

  - Deaths and mid-year population in 2023 (in 1000s)
  
  - One row for each single year of age for every country
  
  - From United Nations World Population Prospects

Start by setting up R and loading the dataset
```{r, message = FALSE, tidy = FALSE}
# Load tidyverse and clear the R environment
library(tidyverse)
rm(list=ls())

# Load dataset and assign it the name country_year_df
country_age_df <- read_csv(url("https://github.com/tomvogl/econ125/raw/main/data/UN_mortality_2023_country_age.csv"))
```

## Variables

Let's look at the first few rows of the dataset
```{r}
head(country_age_df, 3)
```

Demographers use $x$ to denote age, so let's rename `age` as `x` in the data
```{r}
country_age_df <- country_age_df |> rename(x = age)
```

Our building block for today is the age-specific mortality rate at age $x$
$$m_x = 1000 * \frac{d_x}{p_x}$$
where $d_x =$ deaths at age $x$ and $p_x =$ midyear population of age $x$
```{r}
country_age_df <- country_age_df |> mutate(m = 1000*deaths/pop)
```

## Running example: Japan vs Nigeria
Let's split off a data frame with just Japan and Nigeria

```{r}
j_n_df <- country_age_df |> filter(country=="Japan"|country=="Nigeria")
```

Now let's plot the age pattern of mortality by country

```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(j_n_df, aes(x = x, y = m, color = country)) +
  geom_line()
```

Typical rich vs poor comparison: Japanese mortality is lower at every age

Mortality higher in infancy, lower in older childhood, rising in adulthood

## Rescaling y-axis: Japan vs Nigeria
Common to take natural log of mortality rate to see more detail at younger ages

Just write `y = log(m)` instead of `y = m` (equivalent to using a log scale)

```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(j_n_df, aes(x = x, y = log(m), color = country)) +
  geom_line()
```

Many call this pattern the mortality 'swoosh'

Since $\ln(\frac{a}{b}) = \ln(a)-\ln(b)$, gap in $\ln(m)$ informative about ratio of $m$

## Crude mortality rate: definition
Often, we want a single measure to describe the burden of mortality

The simplest measure is the crude mortality rate

- $CMR$ equals total deaths divided by total population

- Low information requirement: only deaths and people, no age

- Commonly applied to settings without death registration systems

- Same as average of age-specific mortality rates, weighted by age distribution
  
  $$CMR = \sum_{x} \text{share}_x * m_x$$
  $\text{share}_x$ is the share of people aged $x$, $m_x$ is the mortality rate at age $x$

## Crude mortality rate: calculation
Let's try it out for Japan and Nigeria
```{r}
j_n_df |>
  group_by(country) |>
  summarise(cmr = sum(pop/sum(pop) * m))
```
In both countries, roughly 12 deaths per 1000 people

Japanese mortality is lower than Nigerian at every age, but $CMR$ is **higher**

What is going on?

## Problem: Japan is much older than Nigeria
Let's calculate age shares and add them to `j_n_df`
```{r}
j_n_df <-
  j_n_df |>
  group_by(country) |>
  mutate(share = pop/sum(pop))
```

Plot them by country, and see the Nigeria has lots more young people
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(j_n_df, aes(x = x, y = share, color = country)) +
  geom_line()
```

## Solution: age-standardized mortality rate
The age-standardized mortality rate ($ASMR$) uses same shares for both countries

Compute $ASMR$ using Nigeria's shares $\rightarrow$ Nigeria $ASMR$ 11x Japan $ASMR$
```{r}
# Step 1: compute Nigeria's shares, save them to a new data frame
nigeria_df <- 
  country_age_df |> 
  filter(country == "Nigeria") |> 
  mutate(nigeria_share = pop / sum(pop)) |> 
  select(x, nigeria_share) # keep these variables

# Step 2: apply Nigeria's shares to both countries, compute ASMR
j_n_df |> 
  left_join(nigeria_df, by = "x") |>
  group_by(country) |> 
  summarise(asmr = sum(nigeria_share * m))
```

## Problem: choice of age structure is arbitrary
Very different answers using Japan's vs Nigeria's age structure: 4x vs 11x
```{r}
# Step 1: compute Japan's shares, save them to a new data frame
japan_df <- 
  country_age_df |> 
  filter(country == "Japan") |> 
  mutate(japan_share = pop / sum(pop)) |> 
  select(x, japan_share)

# Step 2: apply Japan's shares to both countries, compute ASMR
j_n_df |> 
  left_join(japan_df, by = "x") |>
  group_by(country) |> 
  summarise(asmr = sum(japan_share * m))
```

## Solution: life expectancy

In terms of a **hypothetical person**, period life expectancy at age $x$ asks...

*How many more years would a person aged $x$ expect to live if she 
experienced current age-specific mortality rates for the rest of her life?*

\vspace{1cm}
\pause

Can equivalently ask in terms of a **hypothetical group of people**...

*How many more years would a group of people aged $x$ live on average if they
experienced current age-specific mortality rates for the rest of their lives?*

## Functions of the life table

Suppose...

- $l_x$ people are celebrating their $x^{th}$ birthday

- $q_x$ is the probability of dying before the $(x+1)^{th}$ birthday

\pause
Then...

- $d_x = l_x \times q_x$ people die in their $x^{th}$ year

- $l_{x+1} = l_x - d_x = l_x \times (1-q_x)$ survive to the $(x+1)^{th}$ birthday

\pause
Assume that decedents die halfway through the year on average, so...

- $L_x = 0.5 \times d_x + l_{x+1}$ person-years are lived in the $x^{th}$ year

\pause
If we repeat these calculations for ages $x+1$ to $x_{max}$, then...

- $T_x = L_x + L_{x+1} + \cdots + L_{x_{max}}$ person-years are lived above age $x$

And the average person lives...

- $e^o_x = \frac{T_x}{l_x}$ more years after the $x^{th}$ birthday

## Life table

Putting these all together, we obtain a life table

- The initial value of $l_x$ (called the *radix*) is made up and does not affect $e^o_x$

- The death probabilities $q_x$ come from the data

- All other elements are derived from $q_x$ and the initial value of $l_x$

## Deriving $q_x$

The death probability $q_x$ is a little different from the mortality rate $m_x$
$$q_x = \frac{\text{deaths}}{\text{starting population}} \qquad m_x = \frac{\text{deaths}}{\text{midyear population}}$$
Special case of a general demographic principle
$$\text{probability} = \frac{\text{number of occurrences}}{\text{number of trials}} \qquad \text{rate} = \frac{\text{number of occurrences}}{\text{number of person-years lived}}$$
\pause
These definitions are consistent since we assumed deaths halfway through year

- Assumption works well for most ages, though not the very young or very old

- To keep things simple, we will use the halfway assumption for every age

Under the halfway assumption...
$$q_x = \frac{m_x}{1 + 0.5 \times m_x}$$
which adds people who died in the first half if the year back into the denominator

## Calculating $q_x$ in the data

First convert the mortality rate from a 1000 scale to a 1 scale
```{r}
country_age_df <- country_age_df |> mutate(m = m/1000)
```

Now calculate the probability of dying $q_x$
```{r}
country_age_df <- country_age_df |> mutate(q = m / (1 + 0.5 * m))
```

But revise so that we assume everyone at the oldest age (100) dies
```{r}
country_age_df <- 
  country_age_df |> 
  mutate(q = if_else(x==100, 1, q))
```

The function `if_else(A, B, C)` returns `B` if condition `A` is true and `C` otherwise

Here, it replaces `q` with the value 1 if `x` is 100 but leaves it unchanged otherwise

## Plotting $q_x$: Japan vs Nigeria

Let's split off Japan and Nigeria again
```{r}
j_n_df <- country_age_df |> filter(country=="Japan"|country=="Nigeria")
```

Plot $q_x$ over age by country
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}


ggplot(j_n_df, aes(x = x, y = q, color = country)) +
  geom_line()
```

Very similar to $m_x$ except at the oldest age

## Calculating $l_x$ and $d_x$ in the data

Within each country, let's start with a radix ($l_0$) of 100,000 and compute all subsequent $l_x$ and $d_x$

- `arrange()` sorts the data by age 

- `lag()` takes the value of q from the last age

- `default = 1` sets the lagged survival probability to 1 in the first row

- `cumprod()` multiplies together all of the past survival probabilities
```{r}
country_age_df <- 
  country_age_df |> 
  group_by(country) |> 
  arrange(x) |>
  mutate(l = 100000 * cumprod(lag(1-q, default=1)),
         d = l*q)
```

## Plotting $l_x$: Japan vs Nigeria

Split off Japan and Nigeria again
```{r}
j_n_df <- country_age_df |> filter(country=="Japan"|country=="Nigeria")
```

Plot the number of survivors by age, $l_x$
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(j_n_df, aes(x = x, y = l, color = country)) +
  geom_line()
```

Also known as the survival function, common in biomedical research

## Plotting $d_x$: Japan vs Nigeria

Plot the number of deaths by age, $d_x$

```{r, fig.width = 4, fig.height = 2, fig.align = "center"}
ggplot(j_n_df, aes(x = x, y = d, color = country)) +
  geom_line()
```

Lots of elderly deaths in Japan, childhood deaths in Nigeria

This result is due to the shape of $m_x$, not the age structure of the population

## Calculating ${L_x, T_x, e^o_x}$ in the data
Set person-years lived at $x$, $L_x$, to be

- 0.5 for those who die ($d$)

- 1 for those who survive the age ($l-d$)
```{r}
country_age_df <- country_age_df |> mutate(L = 0.5*d + (l-d))
```

\pause
To obtain person-years lived after $x$, $T_x$, for each country...

- Sum person-years lived at all ages using `sum()`

- Subtract person-years lived up through age $x$ using `cumsum()`

- Add back in person-years lived at age $x$
```{r}
country_age_df <- 
  country_age_df |> 
  group_by(country) |>
  arrange(x) |>
  mutate(T = sum(L) - cumsum(L) + L)
```

\pause
To obtain life expectancy $e^o_x$, divide $T_x$ by $l_x$
```{r}
country_age_df <- country_age_df |> mutate(e = T/l)
```

## Plotting $e^o_x$: Japan vs Nigeria

Split off Japan and Nigeria again
```{r}
j_n_df <- country_age_df |> filter(country=="Japan"|country=="Nigeria")
```

Plot life expectancy at each age, $e^o_x$
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(j_n_df, aes(x = x, y = e, color = country)) +
  geom_line()
```

Can easily see the 30-year gap in life expectancy at birth

In Nigeria, $e^o_1 > e^o_0$ due to high infant mortality

## Global distribution of $e^o_0$

Let's plot a histogram of life expectancy **at birth** across countries in 2023

Use `filter()` to create a new data frame that only has rows with $x = 0$
```{r}
country_e0 <- country_age_df |> filter(x==0)
```

- Use `geom_histogram()` from `ggplot()` to plot the histogram
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center", warning = FALSE}
ggplot(country_e0, aes(x = e)) +
  geom_histogram(binwidth=1)
```

## Importance of mortality at younger ages

Life expectancy puts a lot of weight on mortality at younger ages

This is a feature of the measurement tool, not necessarily desirable

To see this point, let's simulate Japanese $e^o_x$ under Nigerian infant mortality

Let's look at actual $e^o_0$ and $q_0$ in Japan and Nigeria...
```{r, echo=FALSE, results="hide"}
j_n_df <- j_n_df |> ungroup()
```

```{r}
j_n_df |> 
  filter(x==0) |>
  select(country, e,q)
```

## Simulating Japanese life expectancy with Nigerian infant mortality

Let's set Japan's probability of dying in the first year to Nigeria's level
```{r}
j_n_df <- 
  j_n_df |> 
  mutate(q = if_else(x==0 & country=="Japan", 0.0715, q))
```

Now let's compute all the other life table functions as before
```{r}
j_n_df <- 
  j_n_df |> 
  group_by(country) |> 
  arrange(x) |>
  mutate(l = 100000 * cumprod(lag(1-q, default=1)),
         d = l*q,
         L = 0.5*d + (l-d),
         T = sum(L) - cumsum(L) + L,
         e = T/l)
```

## Simulation results

The simulation removed 6 years from Japanese life expectancy at birth! 
```{r, echo=FALSE, results="hide"}
j_n_df <- j_n_df |> ungroup()
```

```{r}
j_n_df |> filter(x==0 & country=="Japan") |>  select(e)
```

The reason is that high infant mortality removes survivors from every age > 0
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(j_n_df, aes(x = x, y = l, color = country)) +
  geom_line()
```

## Period vs cohort

Remember that the preceding material all relates to **period** life expectancy

- Asks about a hypothetical group, a.k.a. *synthetic cohort*

- Nobody will ever experience current rates over their lives

- To calculate, only need a snapshot of age-specific mortality in a single year

\vspace{1cm}

Distinct from **cohort** life expectancy

- Asks about an actual group, an actual cohort

- To calculate, need to wait till the whole cohort dies

- Can use same formulas as above or directly compute average age of death