---
title: "Week 4 Methods: Fertility"
subtitle: "ECON 125: The Science of Population"
fontsize: 9pt
output: html_document
---

## Setup

Today, we analyze fertility over time within the United States

We will analyze two separate datasets: first for the full population, and  
second for specific racial and ethnic groups

Our first dataset includes:

  - Births and mid-year female population
  
  - One row for each single year of age (12-55) for each year (1933-2022)
  
  - From the Human Fertility Database, based on US Vital Statistics

Start by setting up R and loading the first dataset
```{r, message = FALSE, tidy = FALSE}
# Load tidyverse and clear the R environment
library(tidyverse)
rm(list=ls())

# Load dataset
age_year_df <- read_csv(url("https://github.com/tomvogl/econ125/raw/main/data/US_births_age_year.csv"))
```

## Variables

Let's look at the first few rows of the dataset
```{r}
head(age_year_df, 3)
```

As we did in the mortality data exercise, let's rename `age` as `x` in the data
```{r}
age_year_df <- age_year_df |> rename(x = age)
```

Our building block for today is the **age-specific fertility rate** at age $x$
$$ASFR_x = 1000 \times \frac{births_x}{women_x}$$
```{r}
age_year_df <- age_year_df |> mutate(asfr = 1000*births/women)
```

## Age-specific fertility rates over time
Let's split off a data frame with every $20^{th}$ year

```{r}
twenty_df <- age_year_df |> filter(year%%20==0)
```

Now let's plot the age pattern of fertility by year
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(twenty_df, aes(x = x, y = asfr, color = factor(year))) +
  geom_line()
```

Age-specific fertility rates are highest in 20s and 30s

Can very easily see the Baby Boom in 1960 and delayed fertility in 2020

## Crude birth rate: definition
As with mortality, we often want a single measure to describe the level of fertility

The simplest measure is the crude birth rate

- $CBR$ equals total births divided by total population (usually $\times 1000$)

$$ CBR = 1000 \times \frac{births}{population} $$

- Low information requirement: only births and people, no age

- Dataset has \# women of childbearing age but not \# people

- So rather than compute $CBR$ ourselves, let's look at a graph from elsewhere

## Crude birth rate: time series
From the World Bank's *World Development Indicators* website

Crude birth rates for the United States (UN data)

## General fertility rate: definition

A slightly more refined measure is the general fertility rate

- $GFR$ equals total births divided by total women of childbearing age (15-44)

$$ GFR = 1000 \times \frac{births}{women \text{ 15-44}} $$

- Slightly higher information requirement than $CBR$: need counts of women by age, but not mothers' ages at birth

- Can calculate in our dataset
```{r}
year_df <-
  age_year_df |>
  mutate(women_childbearing = if_else(x>=15 & x<45, women, 0)) |>
  group_by(year) |>
  summarise(gfr = 1000*sum(births)/sum(women_childbearing))
```

## General fertility rate: time series

Let's look at the time series of the $GFR$
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(year_df, aes(x = year, y = gfr)) +
  geom_line()
```
Compared with the $CBR$ series, the $GFR$ series is:

- Similar after the Baby Boom, with large decline 1960-75 

- Flatter 1990-2010 because GFR omits the growing elderly from denominator

## $CBR$ versus $GFR$

$CBR$ is very susceptible to variation in age structure

- Populations with large elderly shares will tend to have low $CBR$

- $CBR$ still useful $\rightarrow$ directly contributes to the population growth rate

- Recall that in a closed population (no migration), $growth = CBR - CMR$

\vspace{1cm}
\pause

$GFR$ fixes the sensitivity to age structure to some extent

- Removes men, children, and elderly from the denominator

- Still, age structure within 15-44 could vary across populations

- In principle, could calculate age-standardized fertility rate, but uncommon

## Total fertility rate

The more common way to address concerns about age structure is the $TFR$

$TFR$ is a similar thought experiment to period life expectancy

Asks: *If a woman experienced today’s age-specific fertility rates at every age, how many children could she expect to have over her life?*

Despite similarity to period life expectancy, $TFR$ is **much** easier to calculate

Recall the $ASFR$ graph $\rightarrow TFR$ is simply the sum of these rates across all ages 

```{r, echo=FALSE, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(twenty_df, aes(x = x, y = asfr, color = factor(year))) +
  geom_line()
```

## Total fertility rate: definition

With single-year age intervals, the total fertility rate is simply:

$$ TFR = \sum_{x} ASFR_x/1000 $$
Typically, $x$ runs from 15 to 44 or from 15-49

In our data, we have ages 12-55, but with few births before 12 and after 45, so it makes little difference

Calculation in R is simple
```{r}
year_df <-
  age_year_df |>
  group_by(year) |>
  summarise(tfr = sum(asfr/1000))
```

## Total fertility rate: time series

Let's look at the time series of the $TFR$
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(year_df, aes(x = year, y = tfr)) +
  geom_line()
```

Much easier to interpret $TFR$ than $CBR$ and $GFR$! Children per woman...

- 2 in 1930s

- Almost 4 circa 1960 

- At or below 2 since early 1970s

## Thinking about timing

$TFR$ was 3.7 in late 1960 and 1.6 in 2020

Does this mean that women in the 1960s had 3.7 children on average?

No!

\vspace{1cm}
\pause

Timing matters

- Possible for $TFR$ to swing wildly, even with a constant lifetime \# children

- For example, suppose all women have exactly 2 children over their lifetimes

- Suppose gov't announces it will pay \$1m to every pregnant woman this year

- $TFR$ will surge this year and drop next year, but lifetime fertility constant

## Quantum versus tempo

**Quantum effect**  
Refers to **total number of children** a cohort of women have over their lifetimes  
- Example: A drop in completed fertility from 2.1 to 1.8 children per woman

\vspace{1cm}

**Tempo effect**  
Refers to **shifts in the timing** of childbearing  
- Example: Women delay first birth $\rightarrow TFR$ falls but lifetime fertility unchanged
  
  
## Cohort fertility

To learn about the quantum of fertility, need to switch to a cohort perspective

Can approximate cohort fertility by stringing together period age-specific rates

- E.g., $ASFR_{15}^{1990}, ASFR_{16}^{1991}, \cdots, ASFR_{44}^{2019}$

- "Approximate" because $ASFR_{15}^{1990}$ mixes women born in 1974 and 1975

- But still very close to what we would get from following the 1975 cohort

Create age-cohort data frame for ages 15-45

```{r}
age_cohort_df <-
  age_year_df |>
  filter(x>=15 & x<=45) |>
  mutate(birthyear = year - x)
```

## Plot cohort $ASFR_x$ by age

Let's split off a data frame with every $20^{th}$ cohort

```{r}
twenty_df <- age_cohort_df |> filter(birthyear%%20==0)
``` 

And then draw the evolution of age-specific fertility for specific cohorts
```{r, fig.width = 4, fig.height = 2, fig.align = "center"}
ggplot(twenty_df, aes(x = x, y = asfr, color=factor(birthyear))) +
  geom_line()
```

## Translating age-specific rates to children ever born

Let's compute children ever born at age $x$ $(CEB_x)$ for each cohort and age

To do so, we need to focus on cohorts that we can see starting at age 15
```{r}
age_cohort_df <-
  age_cohort_df |>
  group_by(birthyear) |>
  mutate(min_age = min(x)) |>
  filter(min_age==15)
```

And then compute the running sum of age-specific rates
```{r}
age_cohort_df <-
  age_cohort_df |>
  group_by(birthyear) |>
  arrange(x) |>
  mutate(ceb = cumsum(asfr/1000))
```

## Lifecycle children ever born by cohort

Let's split off a data frame with every $20^{th}$ cohort again

```{r}
twenty_df <- age_cohort_df |> filter(birthyear%%20==0)
``` 

And then draw the evolution of $CEB_x$
```{r, fig.width = 4, fig.height = 2, fig.align = "center"}
ggplot(twenty_df, aes(x = x, y = ceb, color=factor(birthyear))) +
  geom_line()
```
1920 and 1940 cohorts had similar lifetime numbers, but 1940 had them earlier

## Completed fertility rate

The completed fertility rate is children ever born at end of reproductive period

$$CFR = CEB_{45}$$

Easy to obtain from our existing data frame: just keep observations with $x=45$
```{r}
cohort_df <-
  age_cohort_df |>
  filter(x==45) |>
  rename(cfr = ceb) |>
  select(birthyear, cfr)
```

What is the distribution of $CFR$ across cohorts?
```{r}
summary(cohort_df)
```
## Plotting completed fertility rate across cohorts

```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(cohort_df, aes(x = birthyear, y = cfr)) +
  geom_line()
```

Can see that women never had more than 3.5 kids on average, nor less than 1.75

Have to wait for cohorts to age out of childbearing $\rightarrow$ cohort patterns out of date

## Back to period data

Explore fertility differences across racial/ethnic groups within the US

Another dataset!

  - Age-specific fertility rates for 5-year age groups (10-14 to 45-49)
  
  - One row for each year (1993-2019) and racial/ethnic group
  
  - From the National Center for Health Statistics, based on US Vital Statistics

```{r, message = FALSE, tidy = FALSE}
# Clear the R environment to keep it uncluttered
rm(list=ls())

# Load dataset
race_year_df <- read_csv(url("https://github.com/tomvogl/econ125/raw/main/data/US_asfr_race_year.csv"))
```

## Variables

Let's look at the first few rows of the dataset
```{r}
head(race_year_df, 3)
```

Different structure from before: now each age group gets its own column

Researchers call this a **wide format** dataset, in contrast with **long format**

## Race/ethnicity

Let's see what race/ethnicity categories are in the data
```{r}
race_year_df |> distinct(race)
```

We want to calculate $TFR$ for each race-year combination

## Teen birth rate by race/ethnicity and year

Interesting to consider the teen birth rate by race/ethnicity over time

Many policies during this period tried to reduce the teen birth rate
```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(race_year_df, aes(x = year, y = asfr1519, color = race)) +
  geom_line()
```
Steep decline in teen fertility, especially for disadvantaged groups

- All groups fell by $>70\%$, largest in absolute terms for Black and Hispanic

## $TFR$ by race/ethnicity and year

Code for calculating $TFR$ will be a bit different from before

- Need to sum across `asfr1014` to `asfr4549`

- Need to multiply by 5 because a woman spends five years in each age group

- No longer need `group_by()` because the observations are already grouped
```{r}
race_year_df <-
  race_year_df |>
  mutate(tfr = 5/1000*(asfr1014 + asfr1519 + asfr2024 + asfr2529 +
                       asfr3034 + asfr3539 + asfr4044 + asfr4549))
```

## Initial $TFR$ by race/ethnicity
When I was a kid in the 1990s, Hispanic, Black, Native women had more kids

Hispanic women prevented the US from below-replacement fertility (<2.1)

We can see this pattern by reporting $TFR$ by race/ethnicity in 1993
```{r}
race_year_df |>
  filter(year==1993) |>
  select(race, tfr) |>
  arrange(race)
```

## $TFR$ time series by race/ethnicity

```{r, fig.width = 4, fig.height = 1.8, fig.align = "center"}
ggplot(race_year_df, aes(x = year, y = tfr, color = race)) +
  geom_line()
```

For a long time, Hispanic women kept US $TFR$ above replacement level

No longer! Hispanic $TFR$ fell 32\% in 2006-19

Native $TFR$ fell 44\% in 1993-2019