---
title: "Week 9 Methods: Time"
subtitle: "ECON 125: The Science of Population"
fontsize: 9pt
output: html_document
---

## Setup

Today, we analyze data on manufacturing plants in Indonesia:

  - Revenue, value added, capital, materials, and workers
  
  - One observation per plant (called a "firm" in the data) per year
  
  - Manufacturing Survey of Large and Medium-Sized Firms (*Statistik Industri*)

Start by setting up R and loading the data
```{r, message = FALSE, tidy = FALSE}
# Load tidyverse and clear the R environment
library(tidyverse)
rm(list=ls())

# Load dataset
load(url("https://github.com/tomvogl/econ125/raw/main/data/indonesia_firms.rds"))

# Ask R to not use scientific notation
options(scipen = 999)
```

## Glimpse

```{r}
glimpse(indonesia_firms)
```

## Years

```{r}
indonesia_firms |> count(year)
```

## Employment

Lots of interesting variables to consider: employemtn, value of materials, value of capital, revenue, value added

We'll focus on the number of workers because it's easy to interpret without worrying about inflation or exchange rates

Can think about this exercise as if we are starting a new manufacturing plant, and we want to know how many workers we should expect to need

The number of workers per plant is very widely dispersed
```{r}
summary(indonesia_firms$workers)
```

## Employment over Time

```{r}
indonesia_firms |> 
  group_by(year) |> 
  summarise(average = mean(workers), 
            p25 = quantile(workers, probs = 0.25), 
            p50 = quantile(workers, probs = 0.50), 
            p75 = quantile(workers, probs = 0.75))
```

## Ages of Plants

The time series of average employment does not give us a good idea of how many workers will be working in our new plant over time

We will need to think about each plant's age: years since opening

Age constructed from observing the same plant ID more than once, so only have it for plants that opened after 1990
```{r}
summary(indonesia_firms$age)
```

Plants present in 1990 have `age = NA`, slightly more than half the observations
```{r}
nrow(indonesia_firms)
```

For the rest of the analysis, drop plants if we don't know their ages
```{r}
indonesia_firms <- indonesia_firms |> drop_na(age)
```

## Ages of Plants

To better understand the structure of the data, let's look at plant age by year
```{r}
table(indonesia_firms$age, indonesia_firms$year)
```

Can you see the consequences of the 1997 Asian financial crisis in the table?

Can you see cohorts in the table?

## Ages and Cohorts of Plants

The dataset also has a cohort variable: the year the plant opened (was "born")

Tabulating age and cohort represents the plant life-cycle in an intuitive way
```{r}
table(indonesia_firms$age, indonesia_firms$cohort)
```

## Cross-Sectional Life-Cycles of Plants

Let's return to thinking about how the number of workers varies with plant age

Let's start naively, by looking at cross-sectional patterns in the final year
```{r}
tbl <-
  indonesia_firms |> 
  filter(year==2000) |> 
  group_by(age) |>
  summarise(average = mean(workers),
            median = median(workers))
```

## Cross-Sectional Life-Cycle Graph

Both average and median are rising, but slopes are different
```{r, fig.width=5, fig.height=2.5}
ggplot(tbl, aes(x=age)) +
  geom_line(aes(y=average, color = "average")) +
  geom_line(aes(y=median, color = "median")) +
  labs(y = "workers")
```

## Cross-Sectional Life-Cycle Graph: Log Scale

Differing slopes may be because employment growth is proportional to size

In that case, the plots would be parallel with a log scale
```{r, fig.width=5, fig.height=2.5}
ggplot(tbl, aes(x=age)) +
  geom_line(aes(y=average, color = "average")) +
  geom_line(aes(y=median, color = "median")) +
  labs(y = "workers") +
  scale_y_log10()
```

## Cross-Sectional Growth Rates

Given the parallel results with a log scale, maybe we should plot growth rates

Let's compute the cross-sectional growth rate of the average and median
```{r}
tbl <-
  tbl |> 
  arrange(age) |> 
  mutate(average_ratio = average / first(average),
         median_ratio = median / first(median))
```

## Cross-Sectional Life-Cycle Graph: Growth Version

Suggests plant size flat for first 5 years, then doubles to triples in next 4
```{r, fig.width=5, fig.height=2.5}
ggplot(tbl, aes(x=age)) +
  geom_line(aes(y=average_ratio, color = "average")) +
  geom_line(aes(y=median_ratio, color = "median")) +
  labs(y = "ratio")
```

## Cohort Life-Cycles of Plants

The cross-sectional patterns may be very misleading

More informative to look at cohorts as they age

For simplicity, we'll focus on averages rather than medians
```{r, message=FALSE}
tbl <-
  indonesia_firms |> 
  group_by(age, cohort) |>
  summarise(average = mean(workers))
```

## Cohort Life-Cycle Graph

```{r, fig.width=5, fig.height=2.5}
ggplot(tbl, aes(x=age, y=average, color=factor(cohort))) +
  geom_line() +
  scale_y_log10()
```
Steady growth with age, but lots of variation in starting points

## Cohort Growth Rates

Growth rates will provide easiest comparison of cohort and cross-section

Let's compute the cohort growth rate of average employment
```{r}
tbl <-
  tbl |> 
  group_by(cohort) |>
  arrange(age) |> 
  mutate(average_ratio = average / first(average)) |>
  ungroup()
```

## Cohort Life-Cycle Graph: Growth Version

The cross-sectional graph showed a 9-year tripling of the average

The cohort graph shows a doubling

The cross-section overstated employment growth because it ignored cohort effects
```{r, fig.width=5, fig.height=2.5}
ggplot(tbl, aes(x=age, y=average_ratio, color=factor(cohort))) +
  geom_line()
```

```{r, echo = FALSE}
# save the plot for later
plot1 <- ggplot(tbl, aes(x=age, y=average_ratio, color=factor(cohort))) +
  geom_line()
```

## Plant Death

Let's look again at our age/cohort table

Do you see plant death?
```{r}
table(indonesia_firms$age, indonesia_firms$cohort)
```

## Plant Survival Function

Can compute survival function, just like in week 2

- Count the number of plants in each age/cohort cell

- Group by cohort

- Compute current number as a share of initial number

- Ungroup
```{r}
tbl <- 
  indonesia_firms |> 
  count(age, cohort) |>
  group_by(cohort) |>
  arrange(age) |> 
  mutate(share_surviving = n / first(n)) |>
  ungroup()
```

## Plant Survival Function

For intuition, let's look at the part of the table for the oldest cohort
```{r}
tbl |> filter(cohort==1991)
```

## Plant Survival Curves

A lot of plants die!

By 9 years, half close down

Alternatively, then median length of operation for a plant is 9 years
```{r, fig.width=5, fig.height=2.5}
ggplot(tbl, aes(x=age, y=share_surviving, color=factor(cohort))) +
  geom_line()
```

## Reconsidering the Cohort Life-Cycle Graph

Now that we have acknowledged that many plants close, we must admit that the cohort graph did **not** track the same plants over the lifecycle

As plants exit the sample, average may change due to selection, not life-cycle
```{r, fig.width=5, fig.height=2.5, echo=FALSE}
plot1
```

## Selection over the Lifecycle

The issue is that cohorts may change in composition over time

Various drivers of this selection

- Here, it is mortality (or plant closing)

- In Abramitzky and Boustan, it is also return migration

- In other studies, it might be variability in who agrees to participate

In repeated cross-section data (as in PS 5), can't do anything about this issue

In panel data (as we have here), can do more, but no perfect solution

## Addressing Selection

One fix for the selection issue: focus on plants that were present in all years

This approach is perfect, since we are not sure our plant is going to survive

Can implement it here by using `years_in_sample`
```{r}
indonesia_firms |> count(years_in_sample)
```

Let's focus on the firms that were present in every year: "balanced panel"

## Employment Growth in the Balanced Panel

Keep the balanced panel, then compute plant-level employment growth
```{r, message=FALSE}
indonesia_firms <-
  indonesia_firms |> 
  filter(years_in_sample==10) |>
  group_by(psid) |>
  arrange(age) |> 
  mutate(ratio = workers / first(workers)) |>
  ungroup()
```

Now compute average ratio
```{r, message=FALSE}
tbl <-
  indonesia_firms |> 
  group_by(age) |>
  summarise(average_ratio = mean(ratio))
```

## Graph of Employment Growth in the Balanced Panel

The graph shows 2.5x employment growth, between the two sets of earlier results

Reflects growth of plants that started in 1991 and kept running through 2000
```{r, fig.width=5, fig.height=2.5}
ggplot(tbl, aes(x=age, y=average_ratio)) +
  geom_line()
```