---
title: "ECON 125 Problem Set 2"
author: "Your name"
output: html_document # can change to pdf_document if you have LaTeX
---

```{r, message = FALSE, include = FALSE}
# This code block sets up the R environment for you
# You do not need to edit it

# Clear the R environment
rm(list=ls())

# Load tidyverse
library(tidyverse)

# Load data into data frame and call it df
df <- read_csv(url("https://github.com/tomvogl/econ125/raw/main/data/US_mortality_age_year.csv"))
```

## Question 1

Question 1 text here.

```{r}
# Question 1 code here
```

\newpage

## Question 2

Question 2 text here.

```{r}
# Question 2 code here
```

\newpage

## Question 3

Question 3 text here.

```{r}
# Question 3 code here
```
\newpage

## Question 4

Question 4 text here.

```{r}
# Question 4 code here
```
\newpage

## Question 5

Question 5 text here.

```{r}
# Question 5 code here
```
\newpage

## Question 6

Question 6 text here.

```{r}
# Question 6 code here
```
\newpage

## Question 7

Question 7 text here.

```{r}
# Question 7 code here
```
\newpage

## Question 8

Question 8 text here.

```{r}
# Question 8 code here
```
\newpage

## Question 9

Question 9 text here.

```{r}
# Question 9 code here
```
\newpage

## Question 10

Question 10 text here.

```{r}
# Question 10 code here
```